Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry

### 2016-11-01 ms_pre_post.Rmd
Minor changes responding to coauthor comments; add more coauthors who will, I hope, agree to be on this paper

---
title: "Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry"
author: "A. Bradley Duthie et al."
date: "August 24, 2016"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Inbreeding avoidance has been widely observed through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no conceptual framework compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: Will evolution of post-copulatory inbreeding avoidance strengthen selection for polyandry? Will selection for a costly inbreeding strategy be negligible given a non-costly alternative strategy? And will selection for an inbreeding strategy be negligible given existence of an already adaptive alternative strategy? We show that selection for polyandry is greatly strengthened by post-copulatory inbreeding avoidance. Strong selection for relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

Meanwhile, empirical studies of diverse species have found evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre-copulatoryor post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre-copulatoryor post-copulatory inbreeding avoidane despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These empirical studies demonstrate that diverse combinations of pre-copulatoryand post-copulatory inbreeding avoidance occur in nature. However, there is as yet no conceptual framework that predicts *a priori* what combinations should be favoured by selection when both pre-copulatoryand post-copulatory inbreeding avoidance, and polyandry can evolve. Such a framework would allow prediction of the form of mating system evolution across different systems that experience inbreeding depression and allow existing empirical studies to be used to test evolutionary predictions through comparative analyses.

<!--- The terminal sentence is non-sensical. It confuses what theory actually is and suggests a post-hoc testing of a model through already existing data -- what does *prediction* even mean in this context !? This is bad science. --->

@@

@@

Will selection for a costly inbreeding strategy be negligible given a non-costly alternative strategy?
----------------------------------------------

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero and could not evolve, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

<!--- Co-author wants to talk about polyandry and stuff: ''But these figures don’t look very symmetrical – there’s stuff going on in Fig 3 that isn't really noted in the text – should we do a bit more with this? The most obvious difference is that polyandry evolves in B but not D - i.e. when pre-cop IA is costly but not when post-cop IA is costly. Wouldn’t it be helpful to point this out and explain it? NB, presumably polyandry is cost-free in all these simulations? Might it be worth re-iterating that too?'' --->

Will selection for a costly inbreeding strategy be negligible given existence of an already adaptive alternative strategy?
----------------------------------------------

@@

<!--- ------------------------------------------------------------------------- --->
<!--- ------------- LEFT OFF HERE --------------------------------------------- --->
<!--- ------------------------------------------------------------------------- --->

>>>>

---
title: "Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry"
author: "A. Bradley Duthie, Greta Bocedi, Ryan Germain, Matthew E. Wolak, and Jane M. Reid"
date: "August 24, 2016"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Inbreeding avoidance has been widely hypothesised to evolve through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no modelling compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: Does evolution of post-copulatory inbreeding avoidance facilitate evolution of costly polyandry? Is evolution of one costly inbreeding strategy inhibited by another non-costly alternative strategy? And is evolution of one inbreeding strategy precluded by existence of another already adaptive strategy. We show that evolution of polyandry is facilitated by post-copulatory inbreeding avoidance. Evolution of relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

<!--- What is 'quantitative theory'? I don't think that I understand what the implied knowledge gap is here, actually. We set up an empirical system to test some interesting hypotheses; I suppose that the knowledge gap is whether or not those specific hypotheses are supported given the parameters we model. The hypothesis testing sort of relates to the brief verbal theory that we mention in the introduction, but this manuscript does not address any theoretical knowledge gap using modelling. I don't think this is an esoteric issue; we have to make it clear that we know what we're doing and why we're doing it. --->

Meanwhile, empirical studies of diverse species have found evidence for female inbreeding avoidance enacted through pre-copulatory or post-copulatory routes, or both. For example, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. Evidence of both pre-copulatory and post-copulatory inbreeding avoidance has accumulated for some well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. While most studies consider either pre-copulatoryor post-copulatory inbreeding avoidance in isolation, a small number of studies consider both. Accordingly, @Liu2014 found evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Other studies found no evidence of pre-copulatoryor post-copulatory inbreeding avoidane despite strong inbreeding depression [e.g., @Reid2014; @Reid2015]. These empirical studies demonstrate that diverse combinations of pre-copulatoryand post-copulatory inbreeding avoidance occur in nature. However, there is as yet no modelling framework that predicts *a priori* what combinations should be favoured by selection when both pre-copulatory and post-copulatory inbreeding avoidance, and polyandry can evolve. Such a framework would allow prediction of the form of mating system evolution across different systems that experience inbreeding depression and allow existing empirical studies to be compared to model predictions.

<!--- I'm sorry, I didn't understand the suggested terminal sentence or the accompanying comment. I'm not sure I'm clear on what we're considering a prediction versus what we're considering a test of prediction, or what 'underlying understanding' refers to in the context of the key hypotheses and where the come from. ''Maybe end on something like this, to emphasise the value of having such a framework (which I think comes both through prediction and through testing predictions and hence underlying understanding?). I expect you’ll want to edit though'' --->

@@

<!--- I am happy to switch back to using statements as titles instead of questions --->

@@

Is evolution of one costly inbreeding strategy inhibited by another non-costly alternative strategy?
----------------------------------------------

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero and could not evolve, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Evolution of post-copulatory inbreeding avoidance faciliated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

Is evolution of one inbreeding strategy precluded by existence of another already adaptive strategy
----------------------------------------------

@@
